Predicting computational reproducibility of data analysis pipelines \\in large population studies using collaborative filtering

### 2018-07-30 paper.tex
Improved description of RFNTL

\paragraph{Complete Columns 
(Figure~\ref{fig:Columns-Sample-Training-set})} The training set is 
sampled by randomly selecting complete columns in the utility matrix, 
that is, complete subject executions. The last selected column might be 
incomplete to meet the exact training ratio. This method respects the 
time constraints. It corresponds to a situation where the collaborative 
filtering method will predict the reproducibility of the subjects in 
the test set from the subjects in the training set.

\paragraph{Complete Rows (Figure~\ref{fig:Rows-Sample-Training-set})} 
The training set is sampled by selecting complete rows in the utility 
matrix, that is, the first files produced by every execution. The last 
selected row might be incomplete to meet the exact training ratio. This 
method respects the time constraints. It corresponds to a situation 
where the processing of all the subjects is launched and interrupted 
before the execution is complete. The collaborative filtering method 
will then predict the reproducibility of the remaining files.

@@

\paragraph{Random Subjects -- RS 
(Figure~\ref{fig:Uniform-S-Sample-Training-set})} This method builds 
the training set by selecting the files from random subjects until the 
training ratio is reached. The file selected in a subject is the file 
with the lowest index in this subject that has not been already 
selected in the training set. This respects the time constraints as files are 
sampled according to their production time-stamps.

\paragraph{Random File Numbers (Uniform) -- RFNU (Figure~\ref{fig:Diagonal-Sample-Training-set})}
The number of files selected for every subject is randomly selected in
a uniform distribution $U(\textit{a},\textit{b})$, where \textit{b} is set to the total
number of files $N_{f}$ and \textit{a} is set according to training ratio $\alpha$ as follows:
\[
  \begin{cases}
          \textit{a} = 0      & \text{if $\alpha \leq 0.5$ }\\

@@

\paragraph{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution $T(a, b, c)$ as in Figure~\ref{fig:triangular}. The mean of the distribution is 
$\frac{a+b+c}{3}$. We set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with two approaches that ensure that the 
average of the distribution is $\alpha N_{f}$: \todo{clarify this part}
\begin{enumerate}
        \item Largest a (RFNTL, Figure~\ref{fig:triangular-L-Sample-Training-set}): a is set to the 
        largest possible value, i.e., b. The average of the 
        distribution is $\frac{2a+N_{f}}{3}$, therefore 
        $a=\frac{3\alpha-1}{2}N_{f}$. When $\alpha \leq 1/3$, the 
        number of files in every subject is sampled from $T(0, 0, N_f)$ 
        with probability $3\alpha$ and set to 0 otherwise, to ensure 
        that the average number of files by subject in the training set 
        is $\alpha N_f$.
        \item Smallest a (RFNTS, Figure~\ref{fig:triangular-S-Sample-Training-set}): a is 
        set to the smallest possible value, i.e., 0. The average of the 
        distribution is $\frac{b+N_{f}}{3}$, therefore 
        $b=N_{f}(3\alpha-1)$. This implies that $\alpha \geq 1/3$, so 
        that $b \geq 0$, and $\alpha \leq 2/3$, so that $b \leq N_f$. 
        When $\alpha < 1/3$, we set $b=0$ and the number of files 
        selected in every subject is sampled from $T(0, 0, N_f)$ with 
        probability $3\alpha$ and set to 0 otherwise, as in RFNTL. 
        When $\alpha \geq 2/3$, we set $b=N_f$ and the number of files 
        selected in every subject is sampled from $T(0, N_f, N_f)$ with 
        probability $3(1-\alpha)$ and set to $N_f$ otherwise, still to 
        ensure that the average number of files by subject in the training set is 
        $\alpha N_f$.
\end{enumerate}
As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNTL method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set, which is not the case for RFNU or RFNTS.
\begin{figure}
\centering
\includegraphics[width=0.5\columnwidth]{figures/triangular.pdf}
\caption{Triangular distribution $T(a, b, c)$}
\label{fig:triangular}
\end{figure}

\paragraph{Random Unreal (Figure 
~\ref{fig:Random-Unreal-Sample-Training-set})} The training set is 
sampled in a random uniform way, regardless of the file generation 
times. This method does not respect the time constraints in 
Equation~\ref{eq:time}, it will be used as a baseline for comparison 
with other methods.

@@

Two experiments have been conducted for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with biases. 
Results will be evaluated using classification accuracy and compared to (1) a dummy classifier that always 
predicts the value in the majority class and (2) the Random Unreal 
method, used as the baseline sampling technique. All reported values are averages
over 5 repetitions. Due to sampling issues, it is possible that the actual 
training ratio obtained with some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true.

@@

\subsubsection{ALS without Bias}

Accuracy results for the different numbers of subject types are 
reported in Figure~\ref{fig:results-synthetic}. By construction, the 
accuracy of the dummy classifier is close to 0.5 for all subject types. 
Random Unreal performs very well for all subject types, which confirms 
that ALS is working correctly. Unsurprisingly, all the other methods 
perform better than the dummy classifier, and their accuracy decreases 
as the number of subject types increases.

@@

Up to 32 subject types, RFNU achieves a 90\% accuracy for a training 
ratio of about 0.85. Therefore, 15\% of the computations could 
potentially be saved in such reproducibility studies.

@@

Results of ALS with bias are reported in 
Figure~\ref{fig:results-synthetic-als-bias}. RS, RFNU and RFNTL remain 
the methods that best compare to Random Unreal, but the achieved 
accuracy is much lower than without Biases. This is presumably due to 
the fact that file biases are all close to 0.5 in the synthetic data. 
In such a situation, biases are detrimental to the accuracy of the 
collaborative filtering process and shouldn't be included.

@@

Results for ALS without bias are reported in 
Figure~\ref{fig:results-real-als}. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets, when the training ratio is larger 
than 0.5. RS and RFNTL are slightly underperforming compared to RFNU. 
Complete rows and complete columns do not even reach the accuracy of 
the dummy classifier. The reference method, Random Unreal, has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.
\todo{why is the performance of columns so variable?}

@@

Results for ALS with bias are reported in 
Figure~\ref{fig:results-real-als-bias}. From a training ratio of 0.5, 
all methods have a very good accuracy for all datasets, except complete 
columns in Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, 
RFNTL and RFNTS also perform well for a training ratio lower than 0.5. 
Overall, the accuracy of the methods is much higher than without bias. 
This is because the file bias is very strong in this dataset, as 
illustrated by the large number of constant lines in the matrices in 
Figure~\ref{fig:utility-matrices}.

@@

Figure~\ref{fig:roc} compares the sampling methods in the ROC space, 
for a training ratio of 0.9, on the synthetic and Freesurfer dataset. 
The PreFreesurfer dataset is not included since specificity of RFNTL, 
RFNU, RS and complete rows is undefined at this training ratio (the 
test set only contains positive elements \todo{check matrix with 
Soudabeh to confirm}). The average sensitivity and specificity values 
are reported in Table~\ref{table:roc}. On average, RFNU is the best 
performing method.

@@

\subsection{Varying the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in ALS, for the RFNU method and the synthetic dataset. We do not find a 
clear effect of the number of factors on the accuracy.
\begin{figure}[h]
\includegraphics[width=\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, RFNU method, synthetic dataset.}
\label{fig:factors}
\end{figure}

\subsection{Error localization}
Figure~\ref{fig:overlap} illustrates the comparison between RFNU, RFNTL 
and RFNTS. The matrices in this Figure represent the union between the 
training and the test set, where the training set is represented in 
black (negatives) and white (positives), and the test set is 
represented in green (true positives), yellow (false negatives), gray 
(true negatives) and red (false positives). This representation 
provides insights regarding where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.

>>>>

\subsubsection{Complete Columns 
(Figure~\ref{fig:Columns-Sample-Training-set})} The training set is 
sampled by randomly selecting complete columns in the utility matrix, 
that is, complete subject executions. The last selected column might be 
incomplete to meet the exact training ratio. This method respects the 
time constraints. It corresponds to a situation where the collaborative 
filtering method will predict the reproducibility of the subjects in 
the test set from the subjects in the training set.

\subsubsection{Complete Rows (Figure~\ref{fig:Rows-Sample-Training-set})} 
The training set is sampled by selecting complete rows in the utility 
matrix, that is, the first files produced by every execution. The last 
selected row might be incomplete to meet the exact training ratio. This 
method respects the time constraints. It corresponds to a situation 
where the processing of all the subjects is launched and interrupted 
before the execution is complete. The collaborative filtering method 
will then predict the reproducibility of the remaining files.

@@

\subsubsection{Random Subjects -- RS 
(Figure~\ref{fig:Uniform-S-Sample-Training-set})} This method builds 
the training set by selecting the files from random subjects until the 
training ratio is reached. The file selected in a subject is the file 
with the lowest index in this subject that has not been already 
selected in the training set. This respects the time constraints as files are 
sampled according to their production time-stamps.

\subsubsection{Random File Numbers (Uniform) -- RFNU (Figure~\ref{fig:Diagonal-Sample-Training-set})}
The number of files selected for every subject is randomly selected in
a uniform distribution $U(\textit{a},\textit{b})$, where \textit{b} is set to the total
number of files $N_{f}$ and \textit{a} is set according to training ratio $\alpha$ as follows:
\[
  \begin{cases}
          \textit{a} = 0      & \text{if $\alpha \leq 0.5$ }\\

@@

\subsubsection{Random File Numbers (Triangular) -- RFNT}
The number of files selected for every subject is randomly selected in
a triangular distribution $T(a, b, c)$ as in Figure~\ref{fig:triangular}. The mean of the distribution is 
$\frac{a+b+c}{3}$. We set \textit{c} to $N_{f}$ and we set 
\textit{a} and \textit{b} with the following two approaches.

\paragraph{Largest a (RFNTL, 
Figure~\ref{fig:triangular-L-Sample-Training-set})} a is set to the 
largest possible value, i.e., b, and b is set accordingly to ensure 
that the average of the distribution is $\alpha N_{f}$. Two cases occur:
\begin{itemize}
\item When $\alpha > 1/3$:
\[ \textit{a}=\textit{b}=\frac{3\alpha-1}{2}N_{f}
\]
\item When $\alpha \leq 1/3$: $a=b=0$, the number of files selected 
in every subject is sampled from $T(0, 0, N_f)$ with probability 
$3\alpha$ and set to 0 otherwise.
\end{itemize}

\paragraph{Smallest a (RFNTS, 
Figure~\ref{fig:triangular-S-Sample-Training-set})} a is set to the 
smallest possible value, i.e., 0, and b is set accordingly to ensure that the average of the
distribution is $\alpha N_{f}$. Three cases occur:
\begin{itemize}
\item When $\alpha < 1/3$: $a=b=0$, the number of files 
selected in every subject is sampled from $T(0, 0, N_f)$ with 
probability $3\alpha$ and set to 0 otherwise.
\item When $1/3 \leq \alpha \leq 2/3$:
\[
          \textit{a} = 0   \quad ; \quad
          \textit{b}=N_{f}(3\alpha-1)
\]
\item When $\alpha \geq 2/3$: $b=N_f$, the number of files selected in 
every subject is sampled from $T(0, N_f, N_f)$ with probability 
$3(1-\alpha)$ 
and set to $N_f$ otherwise.
\end{itemize}

As illustrated in Figures~\ref{fig:triangular-L-Sample-Training-set} 
and~\ref{fig:triangular-S-Sample-Training-set}, the motivation for the 
RFNTL method is to guarantee that, for large enough values of 
$\alpha$, all subjects will have at least a few files in the training 
set (no empty column), which is not the case for RFNU or RFNTS.
\begin{figure}
\centering
\includegraphics[width=0.5\columnwidth]{figures/triangular.pdf}
\caption{Triangular distribution $T(a, b, c)$}
\label{fig:triangular}
\end{figure}

\subsubsection{Random Unreal (Figure 
~\ref{fig:Random-Unreal-Sample-Training-set})} The training set is 
sampled in a random uniform way, regardless of the file generation 
times. This method does not respect the time constraints in 
Equation~\ref{eq:time}, it will be used as a baseline for comparison 
with other methods.

@@

We conducted two experiments for each reproducibility matrix, to evaluate the performance of our 
predictions using (1) ALS without biases, and (2) ALS with biases. We 
compare the performance of our sampling methods to (1) a dummy 
classifier that always predicts the value in the majority class and (2) 
the Random Unreal method, used as the baseline sampling technique. All 
reported values are averages over 5 repetitions. Due to sampling 
issues, it is possible that the actual training ratio obtained with 
some of the sampling methods does not 
match the target one. We checked that the difference between the target 
and real training ratios was lower than 0.01. We used Spark's ALS model 
as available in package \texttt{pyspark.ml}, with 50 factors, a 
regularization parameter of 0.01, a maximum of 5 iterations and non 
negative constraints set to true.

@@

\subsubsection{ALS without Bias}
 Figure~\ref{fig:results-synthetic} shows accuracy results for ALS 
 without bias. By construction, the accuracy of the dummy classifier is 
 close to 0.5 for all subject types. Random Unreal performs very well 
 for all subject types, which confirms that ALS is working correctly. 
 Unsurprisingly, all the other methods 
perform better than the dummy classifier, and their accuracy decreases 
as the number of subject types increases.

@@

@@

Figure~\ref{fig:results-synthetic-als-bias} shows accuracy results for 
ALS with bias. RS, RFNU and RFNTL remain the methods that best compare 
to Random Unreal, but the achieved accuracy is much lower than without 
biases. This is presumably due to 
the fact that file biases are all close to 0.5 in the synthetic data. 
In such a situation, biases are detrimental to the accuracy of the 
collaborative filtering process and shouldn't be included.

@@

Figure~\ref{fig:results-real-als} shows the accuracy results for ALS without bias. Among the methods that performed 
well in the synthetic dataset, RFNU performs the best, with an accuracy 
higher than 0.95 for all datasets, when the training ratio is larger 
than 0.5. RS and RFNTL are slightly underperforming compared to RFNU. 
Complete rows and complete columns do not even reach the accuracy of 
the dummy classifier. The reference method, Random Unreal, has an 
accuracy close to 1, which shows that collaborative filtering works 
well on this dataset too.
\todo{why is the performance of columns so variable?}

@@

Figure~\ref{fig:results-real-als-bias} shows the accuracy results for 
ALS with bias. From a training ratio of 0.5, all methods have a very 
good accuracy for all datasets, except complete columns in 
Figure~\ref{fig:pfs-c5vsc7-bias}. All methods except RFNU, RFNTL and 
RFNTS also perform well for a training ratio lower than 0.5. Overall, 
the accuracy of the methods is much higher than without bias, due to 
the fact that the file bias is very strong in this dataset, as 
it can be seen in Figure~\ref{fig:utility-matrices}.

@@

Figure~\ref{fig:roc} compares the sampling methods in the ROC space, 
for a training ratio of 0.9, on the synthetic and Freesurfer dataset. 
The PreFreesurfer dataset is not included since specificity of RFNTL, 
RFNU, RS and complete rows is undefined at this training ratio (the 
test set only contains positive elements \todo{check matrix with 
Soudabeh to confirm}). The average sensitivity and specificity values 
are reported in Table~\ref{table:roc}, which confirms that RFNU is the best 
performing method on average.

@@

\subsection{Varying the number of factors}
Figure~\ref{fig:factors} shows the effect of the number of factors used 
in the ALS optimization for the RFNU method and the synthetic dataset. We do not find a 
clear effect of the number of factors on the accuracy.
\begin{figure}[h]
\includegraphics[width=\columnwidth]{data/results/means_of_results/ALS/Synthetic/8type-RFNU-Differ-Num-Factors/ALS-8-types-ranks.pdf}
\caption{Effect of the number of factors used in ALS, RFNU method, synthetic dataset.}
\label{fig:factors}
\end{figure}

\subsection{Error localization}

Figure~\ref{fig:overlap} compares the prediction errors made by RFNU, 
RFNTL and RFNTS. The matrices in this Figure represent the training set 
in black (negatives) and white (positives), and the test set in green 
(true positives), yellow (false negatives), gray (true negatives) and 
red (false positives). This representation provides insights regarding 
where, and perhaps why, prediction errors 
occur. At this training ratio, RFNU (Figure~\ref{fig:overlap-rfnu}) 
uniformly samples the number of files per subject between 0.8$N_f$ and 
$N_f$, which enables the training on the last 
files of the pipeline, while maintaining a low number of columns with 
a low training ratio. On the contrary, RFNTL 
(Figure~\ref{fig:overlap-rfntl}) samples the number of files per 
subject from 0.85$N_f$, but the probability to have a complete column 
is 0 (the one complete column in Figure~\ref{fig:overlap-rfntl} is the 
one included to prevent cold start issues), which leads to a ``stripe" 
of errors at the bottom of the matrix. RFNTS 
(Figure~\ref{fig:overlap-rfnts}) doesn't have this issue, as it 
includes many complete columns in the training set. However, it comes at 
the cost of several columns with a number of files lower than 
60\%: in such columns, the prediction is often entirely wrong, which explains
the reduced accuracy compared to RFNU.
